---
output:
  pdf_document: default
  html_document: default
---
# Choosing and evaluating models

## Evaluating models: classification

Building and applying a logistic regression spam model

```{r load, message=FALSE, warning=FALSE}
set.seed(123)
library(tidyverse, warn.conflicts = FALSE)
library(tidymodels, warn.conflicts = FALSE)
```

Using logistic regression to classify emails into spam or non-spam:

```{r listing_5.1}
# reading the file containing spam data
spamD <- readr::read_tsv(
  "https://raw.githubusercontent.com/WinVector/zmPDSwR/master/Spambase/spamD.tsv"
)

# creating training and testing datasets
spamTrain <- dplyr::filter(.data = spamD, rgroup >= 10)
spamTest <- dplyr::filter(.data = spamD, rgroup < 10)

# training the model
spamModel <- stats::glm(
  formula = spam == "spam" ~ .,
  family = stats::binomial(link = "logit"),
  data = dplyr::select(spamTrain, -rgroup)
)

# looking at the result
broom::tidy(spamModel)

# looking at the model summary
broom::glance(spamModel)

# with predicted response on training data
spamTrain <- broom::augment(
  x = spamModel,
  newdata = spamTrain,
  type.predict = "response"
)

# with predicted response on test data
spamTest <- broom::augment(
  x = spamModel,
  newdata = spamTest,
  type.predict = "response"
)

# performance with the training data
train_perform <- table(y = spamTrain$spam, glmPred = spamTrain$.fitted > 0.5)
colnames(train_perform) <- c("non-spam", "spam")

# looking at performance measures
caret::confusionMatrix(train_perform)

# performance with the test data
test_perform <- table(y = spamTest$spam, glmPred = spamTest$.fitted > 0.5)
```

Looking at actual and predicted sample responses

```{r listing_5.2}
sample <- spamTest[c(7, 35, 224, 327), c("spam", ".fitted")]
print(sample)
```

Spam confusion matrix (to assess performance of the model)

```{r listing_5.3}
# performance with the test data
(cM <- table(truth = spamTest$spam, prediction = spamTest$.fitted > 0.5))
```

Assessing the performance

```{r new_5.1}
# chaning column names to align with rownames
colnames(cM) <- c("non-spam", "spam")

# looking at performance measures
caret::confusionMatrix(cM)
```

Entering data by hand (example of a good spam filter at WinVectorLLC blog)

```{r listing_5.4}
t <- as.table(matrix(data = c(288 - 1, 17, 1, 13882 - 17), nrow = 2, ncol = 2))

rownames(t) <- rownames(cM)
colnames(t) <- c("non-spam", "spam")

# looking at performance measures
caret::confusionMatrix(t)
```

Accuracy for our `glm`-based classifier was 92.14%, while it's 99.87% for a good
classifier.

## Evaluating models: scoring methods

Plotting residuals for evaluating scoring methods

```{r listing_5.5}
# making a dataframe
d <- data.frame(y = (1:10)^2, x = 1:10)

# augmented dataframe for a linear model
d <- broom::augment(
  x = stats::lm(y ~ x, data = d),
  newdata = d
)

# plot
ggplot(data = d) +
  geom_point(aes(x = x, y = y)) +
  geom_line(aes(x = x, y = .fitted), color = "blue") +
  geom_segment(aes(x = x, y = .fitted, yend = y, xend = x)) +
  scale_y_continuous("")
```

## Evaluating models: probability estimation

Double density plot for evaluating probability methods

```{r listing_5.6}
ggplot(data = spamTest) +
  geom_density(aes(x = .fitted, color = spam, linetype = spam))
```

Plotting the receiver operating characteristic (ROC) curve for evaluating
probability methods

```{r listing_5.7}
suppressPackageStartupMessages(library(pROC))

# AUC
(roc_object <- pROC::roc(response = spamTest$spam, predictor = spamTest$.fitted))

# plot with ggplot
pROC::ggroc(roc_object) +
  labs(
    title = "ROC curve",
    x = "False Positive Rate (1-Specificity)",
    y = "True Positive Rate (Sensitivity)"
  )
```

Calculating log likelihood

```{r listing_5.8}
# model log likelihood the model assigns to the test data
sum(base::ifelse(
  test = spamTest$spam == "spam",
  yes = log(spamTest$.fitted),
  no = log(1 - spamTest$.fitted)
))

# log likelihood rescaled by the number of data points
sum(base::ifelse(
  test = spamTest$spam == "spam",
  yes = log(spamTest$.fitted),
  no = log(1 - spamTest$.fitted)
)) / dim(spamTest)[[1]]
```

Computing the null model’s log likelihood
(if the model is a good explanation, then the data should
look likely (not implausible) under the model)

```{r listing_5.9}
# null model
pNull <-
  sum(base::ifelse(
    test = spamTest$spam == "spam",
    yes = 1,
    no = 0
  )) / dim(spamTest)[[1]]

# null model LL
sum(base::ifelse(
  test = spamTest$spam == "spam",
  yes = 1,
  no = 0
)) * log(pNull) +
  sum(base::ifelse(
    test = spamTest$spam == "spam",
    yes = 0,
    no = 1
  )) * log(1 - pNull)
```

Calculating entropy and conditional entropy

```{r listing_5.10, message = FALSE, warning = FALSE}
suppressPackageStartupMessages(library(DescTools))

DescTools::Entropy(x = table(spamTest$spam))

# custom function to compute conditional entropy from the book
conditionalEntropy <- function(t) {
  (sum(t[, 1]) * DescTools::Entropy(t[, 1]) +
    sum(t[, 2]) * DescTools::Entropy(t[, 2])) / sum(t)
}

print(conditionalEntropy(cM))

# package function to do the same
infotheo::condentropy(
  X = spamTest$.fitted > 0.5,
  Y = spamTest$spam,
  method = "emp"
)

# not sure why the different results
```

## Evaluating models: clustering

Clustering random data in the plane

```{r listing_5.11}
set.seed(32297)
d <- tibble::as_tibble(data.frame(x = runif(100), y = runif(100)))
clus <- stats::kmeans(x = d, centers = 5)

# getting cluster assignment as a column (currently, this doesn't work)
# d <- broom::augment(x = clus, data = d)

# cluster assignment column
d$cluster <- clus$cluster

# calculating the size of each cluster
table(d$cluster)
```

Plotting our clusters

```{r listing_5.12}
library(grDevices)

# custom function for computing coordinates for drawing polygons
# `chull` function computes the subset of points which lie on the convex hull of
# the set of points specified
poly_coords <- function(c) {
  f <- subset(d, cluster == c)
  f[grDevices::chull(f), ]
}

# dataframe with coordinates
h <- purrr::map_dfr(.x = unique(clus$cluster), 
                    .f = ~ poly_coords(c = .))

# plot
ggplot() +
  geom_text(data = d,
            aes(
              label = cluster,
              x = x,
              y = y,
              color = cluster
            ),
            size = 3) +
  geom_polygon(
    data = h,
    aes(
      x = x,
      y = y,
      group = cluster,
      fill = as.factor(cluster)
    ),
    alpha = 0.4,
    linetype = 0
  ) +
  theme(legend.position = "none")
```

Calculating the typical distance between items in every pair of clusters

```{r listing_5.14}
# library(reshape2)
# n <- dim(d)[[1]]
# pairs <- data.frame(
#   ca = as.vector(outer(1:n, 1:n, function(a, b) d[a, "cluster"])),
#   cb = as.vector(outer(1:n, 1:n, function(a, b) d[b, "cluster"])),
#   dist = as.vector(outer(1:n, 1:n, function(a, b)
#     sqrt((d[a, "x"] - d[b, "x"])^2 + (d[a, "y"] - d[b, "y"])^2)))
# )
# 
# dcast(pairs, ca ~ cb, value.var = "dist", mean)
```
